Multi-scale assessment of a grassland productivity model

### 2021-02-19 manuscript/manuscript.Rmd
minor equation corrections
Revision round2 (#14)
* minor edits from 2nd biogeosciences revision

* latest pdf

Here we evaluate a low-dimensional model with the intention of it driving productivity projections. The PhenoGrass model developed by @hufkens2016 is a pulse-response productivity model with temperature and precipitation as the primary drivers. The model is parameterized using observations from the PhenoCam network, which have a small spatial resolution  (footprints of < 1ha), sub-daily sampling and sites across all major biomes. These attributes make the PhenoGrass model potentially widely applicable. We expand on the evaluation of the original study by using 84 PhenoCam sites, totalling 89 distinct time series, with 463 site-years of data. We test the model’s performance across varying combinations of North American ecoregions and vegetation types to find an optimal spatial scale in which to parameterize and apply the model. Finally we address where the model performs poorly and how productivity projections for these areas could be implemented or improved.

@@

For historic precipitation and temperature we used the 1-km resolution Daymet dataset [@daymet], extracting daily time series for the pixel at each PhenoCam tower location. Daily mean temperature was calculated as the average between the Daymet daily minimum and maximum temperature, and smoothed with a 15 day moving average. Potential evapotranspiration was calculated using the Hargreaves equation [@hargreaves1985]. Soil wilting point and field capacity were extracted at each Phenocam location from a global dataset [@soil2000].

@@

\begin{equation}
NSE_{j} = 1- \frac{\sum_{i=1}^{n}(fCover_{obs} - fCover_{pred})^{2}}{\sum_{i=1}^{n}(fCover_{obs} - \overline{fCover_{obs}})^{2}}
\end{equation}

@@

@hufkens2016 originally evaluated the PhenoGrass model using 14 grassland sites distributed among seven North American ecoregions. In their evaluation they had an average R\textsuperscript{2} of 0.71, while here the model performed poorly when using more than one ecoregion. It’s likely that the original 14 grassland sites were ideal locations for the PhenoGrass model, since on average they have a single greenup season every year in the spring or summer (Fig. 5A). The additional 24 grassland sites used in the current study have high seasonal variability and elongated growing seasons (Fig. 5B, Fig. S1), and were thus more difficult to represent in a single continental scale grassland model. This highlights the need for longer time series in evaluating low-dimensional models as it may take many years for a single location to experience the full range of variability. As the PhenoCam data archive grows, temporally out of sample validation can be done to better evaluate performance into novel conditions.

>>>>

Here we evaluate a low-dimensional model with the intention of it driving productivity projections. The PhenoGrass model developed by @hufkens2016 is a pulse-response productivity model with temperature and precipitation as the primary drivers. The model is parameterized using observations from the PhenoCam network, which have a small spatial resolution  (footprints of < 1ha), sub-daily sampling and sites across all major biomes. These attributes make the PhenoGrass model potentially widely applicable. We expand on the evaluation of the original study by using 84 PhenoCam sites, totalling 89 distinct time series, with 463 site-years of data. Using the original methods and input datasets we test the model’s performance with this expanded dataset. We also evaluate the model across different combinations of North American ecoregions and vegetation types to find an optimal spatial scale in which to parameterize and apply the model. Finally we address where the model performs poorly and how productivity projections for these areas could be implemented or improved.

@@

For historic precipitation and temperature we used the 1-km resolution Daymet dataset [@daymet], extracting daily time series for the pixel at each PhenoCam tower location. Daily mean temperature was calculated as the average between the Daymet daily minimum and maximum temperature, and smoothed with a 15 day moving average. Potential evapotranspiration was calculated using the Hargreaves equation [@hargreaves1985]. Soil wilting point and field capacity were extracted at each Phenocam location from a global dataset [@soil2000]. Despite other options being available we chose to use the same climate and soil datasets as those used in @hufkens2016 so that any differences in results can be attributed to the expanded PhenoCam dataset used here.

@@

\begin{equation}
NSE_{j} = 1- \frac{\sum_{i=1}^{n}(fCover_{i,obs} - fCover_{i,pred})^{2}}{\sum_{i=1}^{n}(fCover_{i,obs} - \overline{fCover_{obs}})^{2}}
\end{equation}

@@

@hufkens2016 originally evaluated the PhenoGrass model using 14 grassland sites distributed among seven North American ecoregions. In their evaluation they had an average NSE of 0.71, while here the model performed poorly when using more than one ecoregion. It’s likely that the original 14 grassland sites were ideal locations for the PhenoGrass model, since on average they have a single greenup season every year in the spring or summer (Fig. 5A). The additional 24 grassland sites used in the current study have high seasonal variability and elongated growing seasons (Fig. 5B, Fig. S1), and were thus more difficult to represent in a single continental scale grassland model. This highlights the need for longer time series in evaluating low-dimensional models as it may take many years for a single location to experience the full range of variability. As the PhenoCam data archive grows, temporally out of sample validation can be done to better evaluate performance into novel conditions. Further work could use the expanded PhenoCam timeseries to compare with flux tower derived productivity, which @hufkens2016 found was well correlated at the five original locations with available flux data.
